DCMIP2016, Part 1: Models and Equation Sets

### 2016-12-22 DCMIP2016-Part1-Models.tex
- Substantial updates to write-up

{\color{red}INSTRUCTIONS FOR AUTHORS: Fill in text in section 3, 4, 5, 6, 7 and 8 below.}

@@

Several methods are available for subdividing the triangular regions.  One such approach is implemented by the ICON grid generator, which allows an ``arbitrary'' subdivision factor $n$ for the first refinement step only, the so-called root refinement. Typical choices are $n$ = 2, 3 or 5. All additional $m$ refinement steps use $n$ = 2, i.e. are bisection steps. A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices. The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001}.  For a discussion of the effective resolution see \cite{dipankar2015large}.  The ICON grid generator further allows for inset regional grids, produced by additional refinement steps that are only applied over a limited region, or set of regions.  The dynamical core then allows for either one-way or two-way coupling of the refined region to the parent model. The current operational numerical weather prediction of the Deutscher Wetterdienst (German Weather Service, DWD) for instance uses a $R3B7$ global grid with 2949120 cells and 13 km mean resolution in combination with a refined region over Europe at 6.5 km resolution.

@@

@@

Further prognostic relationships can be derived from the thermodynamic equation, including the diabatic heating rate $J$,
\begin{align}
\frac{1}{T} \diff{T}{t} - \frac{\kappa}{p} \diff{p}{t} =& \frac{J}{c_p^\ast},
\end{align} which can be alternatively written as
\begin{align}
\diff{\theta}{t} = \frac{J \theta}{c_p^\ast}, \quad \mbox{or} \quad \diff{\theta_v}{t} = \frac{J \theta_v}{c_p^\ast}.
\end{align}  These equations can then be combined with (\ref{eq:ContinuityEquation}) to obtain
\begin{align}
\pdiff{}{t} (\rho \theta_v) + \nabla \cdot (\rho \theta_v \vb{u}) = \frac{J \rho \theta_v}{c_p^\ast},
\end{align} or similarly for $\theta$.  In conjunction with the material derivative of the ideal gas law,  
\begin{align}
\frac{1}{p} \diff{p}{t} = \frac{1}{\rho} \diff{\rho}{t} + \frac{1}{T_v} \diff{T_v}{t},
\end{align} the thermodynamic equation can be written in the form
\begin{align}
\frac{c_v^\ast}{R^\ast T_v} \diff{T_v}{t} - \frac{1}{\rho} \diff{\rho}{t} = \frac{J}{R^\ast}.
\end{align}  Then substituting (\ref{eq:ContinuityEquation}) gives a prognostic equation for virtual temperature,
\begin{align}
\frac{c_v^\ast}{R^\ast} \diff{T_v}{t} + T_v \nabla \cdot \vb{u} = \frac{J T_v}{R^\ast}.
\end{align}  The prognostic equation for temperature is identical except with $T$ substituted for $T_v$.  An analogous equations for pressure can be obtained through a similar procedure, {\color{red}[Double check these two]}
\begin{align} \label{eq:PressurePrognosticEq}
\frac{c_v^\ast}{c_p^\ast} \diff{p}{t} + p \nabla \cdot \vb{u} = \frac{J p}{c_p^\ast}.
\end{align}  And similarly for Exner pressure,
\begin{align}
\frac{c_v^\ast}{R^\ast} \diff{\pi}{t} + \pi \nabla \cdot \vb{u} = \frac{J \pi}{c_p^\ast}.
\end{align}

@@

In coordinate-invariant form the prognostic velocity equations may be written in either the Lagrangian or Eulerian frame as
\begin{align} \label{eq:PrognosticVelocity1}
\diff{\vb{u}}{t} = \pdiff{\vb{u}}{t} + \vb{u} \cdot \nabla \vb{u} = - \frac{1}{\rho} \nabla p - 2 \vg{\Omega} \times \vb{u} - \nabla \Phi,
\end{align} where $\vg{\Omega}$ denotes the planetary vorticity vector and $\Phi$ is the geopotential function.  The three terms on the right-hand-side of this expression correspond to pressure gradient, Coriolis, and gravitational force, respectively.  In Eulerian form one must be careful with the treatment of the momentum advection term $\vb{u} \cdot \nabla \vb{u}$, since in an arbitrary coordinate frame this term will give rise to Christoffel symbols associated with derivatives of the vector basis {\color{red}[citation needed]}.  Note that it is common to rewrite the pressure gradient force using the relationship
\begin{align}
- \frac{1}{\rho} \nabla p = - c_p^\ast \theta_v \nabla \pi,
\end{align} which follows from (\ref{eq:DiagnosticRelationships1}).  A second form of (\ref{eq:PrognosticVelocity1}) emerges on substituting the vector calculus identity
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \nabla K + \vg{\zeta} \times \vb{u},
\end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D Kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.  This gives rise to the vector-invariant form,
\begin{align} \label{eq:PrognosticVelocity2}
\pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - \nabla (K + \Phi) - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}. 
\end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis that would emerge in (\ref{eq:PrognosticVelocity1}).  In conjunction with (\ref{eq:ContinuityEquation}), both (\ref{eq:PrognosticVelocity1}) also give rise to the flux-form momentum equations,
\begin{align}
\pdiff{}{t} (\rho \vb{u}) = - \nabla \cdot (\vb{u} \otimes \vb{u} + \mathcal{I} p) - 2 \vg{\Omega} \times (\rho \vb{u}) - \rho \nabla \Phi,
\end{align} where $\vb{u} \otimes \vb{u}$ denotes the outer product and $\mathcal{I}$ is the identity matrix.

@@

Due to its association with hydrostatic models, it is common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.  Decomposing the momentum transport term into horizontal and vertical components gives
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w) \vb{k}.
\end{align}  The first term in this expression admits the relationships
\begin{align}
\left[ \vb{u}_h \cdot \nabla \vb{u}_h \right]_z =& \nabla_z K_2 + \zeta_h \vb{k} \times \vb{u}_h, \\
(\vb{u}_h \cdot \nabla \vb{u}_h) \cdot \vb{k} =& - \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) = - K_2 (\nabla \cdot \vb{k}) - \tfrac{1}{2} \left[ \vb{u}_h \cdot (\nabla \times \vb{u}_t)  + (\nabla \times \vb{u}_h) \cdot \vb{u}_t \right]
\end{align} where $\zeta_h = (\nabla \times \vb{u}_h) \cdot \vb{k}$ is the relative vorticity scalar and $\vb{u}_t = \vb{k} \times \vb{u}_h$.  Note that this equation does incorporate metric terms associated with horizontal advection of $\vb{k}$ which must be accounted for.

@@

The dynamics equations are then formulated in terms of the vertical coordinate $s(t,x,y,z)$ with $\partial s / \partial z \neq 0$ everywhere.  In the new coordinate only $s$ is a function of $z$, and so by chain rule we have
\begin{align} \label{eq:VerticalChainRule}
\pdiff{}{z} = \pdiff{s}{z} \pdiff{}{s}.
\end{align}  Chain rule also motivates the definition of the horizontal gradient operator on $s$ surfaces, denoted $\nabla_s$, 
\begin{align}
\nabla_s \equiv \nabla_z - \left( \pdiff{s}{z} \right)^{-1} ( \nabla_z s ) \pdiff{}{z}.
\end{align}  Applying this operator to $z$ gives
\begin{align}
\nabla_z s = - \left( \pdiff{s}{z} \right) \nabla_s z,
\end{align} which in conjunction with (\ref{eq:VerticalChainRule}) admits the more commonly observed relationship
\begin{align} \label{eq:HorizontalChainRule}
\nabla_z = \nabla_s - \pdiff{s}{z} (\nabla_s z) \pdiff{}{s}.
\end{align}  This equation can be used directly to rewrite (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2}) in terms of derivatives over $s$.  Note that the operators $\nabla_z$ and $\nabla_s$ are usually introduced in the context of 2D flows, however the construction described here has the advantage of working seamlessly in a 3D context, while admitting the properties $\vb{k} \cdot \nabla_z = 0$ and $\vb{k} \cdot \nabla_s = 0$.

The generalized velocity $\dot{s}$ following a fluid parcel is defined by
\begin{align} \label{eq:DefinitionDotS}
\dot{s} \equiv \diff{s}{t} = \pdiff{s}{t} + \vb{u} \cdot \nabla s = \pdiff{s}{t} + \vb{u}_h \cdot \nabla_z s + w \pdiff{s}{z}.
\end{align}  A diagnostic relationship for $w$ in terms of $\dot{s}$ then follows,
\begin{align}
w = \left( \pdiff{s}{z} \right)^{-1} \left( \dot{s} - \pdiff{s}{t} - \vb{u}_h \cdot \nabla_z s \right).
\end{align} Substituting this expression back into (\ref{eq:MaterialDerivativeScalar}) gives the material derivative of scalars on $s$ surfaces,
\begin{align}
\diff{}{t} =& \pdiff{}{t} + \vb{u}_h \cdot \left( \nabla_z - \left( \pdiff{s}{z} \right)^{-1} (\nabla_z s) \pdiff{}{z} \right) + \left( \dot{s} - \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} \pdiff{}{z}, \nonumber \\
=& \pdiff{}{t} + \vb{u}_h \cdot \nabla_s + \left( \dot{s} - \pdiff{s}{t} \right) \pdiff{}{s}.
\end{align}

@@

In conjunction with (\ref{eq:HorizontalChainRule}), the horizontal momentum equation (either  (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2})) with an arbitrary vertical coordinate gives rise to a two-term pressure gradient.  This can be avoided by prognosing the covariant components of the velocity in place of the physical velocity components.  The horizontal covariance operator is defined by
\begin{align}
[\vb{b}]_s \equiv [\vb{b}]_z + (\nabla_s z) (\vb{k} \cdot \vb{b}).
\end{align} Hence the covariant horizontal velocity is given by
\begin{align}
\vb{v}_h \equiv [\vb{u}]_s = \vb{u}_h + (\nabla_s z) w.
\end{align}  For a time-dependent $s$ coordinate,
\begin{align}
\pdiff{}{t} (\nabla_s z) =& \pdiff{}{t} \left[ - (\nabla_z s) \left( \pdiff{s}{z} \right)^{-1} \right] \\
=& - \left( \nabla_z \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} + (\nabla_z s) \left( \pdiff{s}{z} \right)^{-2} \pdiff{}{z} \pdiff{s}{t} \\
=& - \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right)
\end{align}  And so the evolution equation for the covariant velocity is
\begin{align}
\pdiff{\vb{v}_h}{t} =& \pdiff{\vb{u}_h}{t} + \pdiff{w}{t} (\nabla_s z) + w \pdiff{}{t} (\nabla_s z) \\
=& - \frac{1}{\rho} \nabla_s p - \nabla_s (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_s - w \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right).
\end{align}  Although a time-dependent $s$ is admitted by this equation, its fully Eulerian nature makes it difficult to use in this case since degrees of freedom are not guaranteed to align with constant $s$ surfaces.

\subsection{Floating Lagrangian Formulation}

Under the floating Lagrangian formulation, the coordinate $s$ is chosen so as to satisfy $\dot{s} = 0$.  This can be accomplished by choosing $s$ to equal potential temperature, although this has issues with monotonicity, or an imaginary conservative tracer \citep{L2004MWR} (for instance, the altitude of the fluid parcel at its initial time).

A consequence of the use of floating Lagrangian coordinates is the need for vertical remapping to avoid severe deformation of coordinate surfaces.  Several options are available, with the piecewise-parabolic method being perhaps the most widely adopted \citep{colella1984piecewise, L2004MWR}.

In the case of time-varying $s$ coordinates, degrees of freedom are usually stored on $s$ surfaces.  To track floating surfaces of constant $s$ we define the floating Lagrangian derivative,
\begin{align}
\frac{d_s}{dt} = \pdiff{}{t} + w \vb{k} \cdot \nabla,
\end{align} which is used to define the tendency in variables that are allowed to ``float'' vertically along surfaces that adjust with the advection velocity.  From (\ref{eq:VerticalVelocityOrthogonal}), the vertical velocity equation then takes the form
\begin{align}
\frac{d_s w}{d t} =& \pdiff{w}{t} + w \pdiff{w}{z} \\
=& - \frac{1}{\rho} \left( \pdiff{s}{z} \right) \pdiff{p}{s} - \left( \pdiff{s}{z} \right) \pdiff{}{s} (K_2 + \Phi) - [ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} ] \cdot \vb{k}.
\end{align}  From (\ref{eq:PrognosticUhEquationK2}), the horizontal velocity equation takes the form
\begin{align}
\frac{d_s \vb{u}_h}{d t} =& \pdiff{\vb{u}_h}{t} + w \vb{k} \cdot \nabla \vb{u}_h \\
=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}),
\end{align} where we have also used
\begin{align}
w \vb{k} \cdot \nabla \vb{u}_h = - w \vb{u}_h \cdot \nabla \vb{k} - (w \vb{k}) \times (\nabla \times \vb{u}_h).
\end{align}  Note that under the shallow atmosphere approximation, the $(\vb{u}_h \cdot \nabla \vb{k})$ term is typically dropped.

>>>>

%{\color{red}INSTRUCTIONS FOR AUTHORS: Fill in text in section 3, 4, 5, 6, 7 and 8 below.}

@@

Several methods are available for subdividing the triangular regions.  One such approach is implemented by the ICON grid generator, which allows an ``arbitrary'' subdivision factor $n$ for the first refinement step only, the so-called root refinement. Typical choices are $n$ = 2, 3 or 5. All additional $m$ refinement steps use $n$ = 2, i.e. are bisection steps. A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices. The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001}.  An example of such a grid is depicted in Figure \ref{fig:ICONMesh}.  For a discussion of the effective resolution see \cite{dipankar2015large}.  The ICON grid generator further allows for inset regional grids, produced by additional refinement steps that are only applied over a limited region, or set of regions.  The dynamical core then allows for either one-way or two-way coupling of the refined region to the parent model. The current operational numerical weather prediction of the Deutscher Wetterdienst (German Weather Service, DWD) for instance uses a $R3B7$ global grid with 2949120 cells and 13 km mean resolution in combination with a refined region over Europe at 6.5 km resolution.

\begin{figure}
\begin{center}
\includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_nested_schematic.pdf}
\end{center}
\caption{An icosahedral (triangular) grid with additional refinement over Europe (shown in red).} \label{fig:ICONMesh}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_dual_R2B3_noWorldMap.pdf}
\end{center}
\caption{An icosahedral (hexagonal) grid.} \label{fig:ICONIcosahedlalHexGrid}
\end{figure}

@@

Further prognostic relationships can be derived from the thermodynamic equation, including the diabatic heating rate $J$,
\begin{align}
\frac{1}{T} \diff{T}{t} - \frac{\kappa}{p} \diff{p}{t} =& \frac{J}{c_p^\ast},
\end{align} which can be alternatively written as
\begin{align}
\diff{\theta}{t} = \frac{J \theta}{c_p^\ast}, \quad \mbox{or} \quad \diff{\theta_v}{t} = \frac{J \theta_v}{c_p^\ast}.
\end{align}  These equations can then be combined with (\ref{eq:ContinuityEquation}) to obtain
\begin{align}
\pdiff{}{t} (\rho \theta_v) + \nabla \cdot (\rho \theta_v \vb{u}) = \frac{J \rho \theta_v}{c_p^\ast},
\end{align} or similarly for $\theta$.  In conjunction with the material derivative of the ideal gas law,  
\begin{align}
\frac{1}{p} \diff{p}{t} = \frac{1}{\rho} \diff{\rho}{t} + \frac{1}{T_v} \diff{T_v}{t},
\end{align} the thermodynamic equation can be written in the form
\begin{align}
\frac{c_v^\ast}{R^\ast T_v} \diff{T_v}{t} - \frac{1}{\rho} \diff{\rho}{t} = \frac{J}{R^\ast}.
\end{align}  Then substituting (\ref{eq:ContinuityEquation}) gives a prognostic equation for virtual temperature,
\begin{align}
\frac{c_v^\ast}{R^\ast} \diff{T_v}{t} + T_v \nabla \cdot \vb{u} = \frac{J T_v}{R^\ast}.
\end{align}  The prognostic equation for temperature is identical except with $T$ substituted for $T_v$.  An analogous equations for pressure can be obtained through a similar procedure,
\begin{align} \label{eq:PressurePrognosticEq}
\frac{c_v^\ast}{c_p^\ast} \diff{p}{t} + p \nabla \cdot \vb{u} = \frac{J p}{c_p^\ast}.
\end{align}  And similarly for Exner pressure,
\begin{align}
\frac{c_v^\ast}{R^\ast} \diff{\pi}{t} + \pi \nabla \cdot \vb{u} = \frac{J \pi}{c_p^\ast}.
\end{align}

@@

In coordinate-invariant form the prognostic velocity equations may be written in either the Lagrangian or Eulerian frame as
\begin{align} \label{eq:PrognosticVelocity1}
\diff{\vb{u}}{t} = \pdiff{\vb{u}}{t} + \vb{u} \cdot \nabla \vb{u} = - \frac{1}{\rho} \nabla p - 2 \vg{\Omega} \times \vb{u} - \nabla \Phi,
\end{align} where $\vg{\Omega}$ denotes the planetary vorticity vector and $\Phi$ is the geopotential function.  The three terms on the right-hand-side of this expression correspond to pressure gradient, Coriolis, and gravitational force, respectively.  In Eulerian form one must be careful with the treatment of the momentum advection term $\vb{u} \cdot \nabla \vb{u}$, since in an arbitrary coordinate frame this term will give rise to Christoffel symbols associated with derivatives of the vector basis {\color{red}[citation needed]}.  Note that it is common to rewrite the pressure gradient force using the relationship
\begin{align}
- \frac{1}{\rho} \nabla p = - \frac{R_d c_p^\ast}{R^\ast} \theta_v \nabla \pi,
\end{align} which follows from (\ref{eq:DiagnosticRelationships1}).  A second form of (\ref{eq:PrognosticVelocity1}) emerges on substituting the vector calculus identity
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \nabla K + \vg{\zeta} \times \vb{u},
\end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D Kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.  This gives rise to the vector-invariant form,
\begin{align} \label{eq:PrognosticVelocity2}
\pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - \nabla (K + \Phi) - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}. 
\end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis that would emerge in (\ref{eq:PrognosticVelocity1}).  In conjunction with (\ref{eq:ContinuityEquation}), both (\ref{eq:PrognosticVelocity1}) also give rise to the flux-form momentum equations,
\begin{align}
\pdiff{}{t} (\rho \vb{u}) = - \nabla \cdot (\vb{u} \otimes \vb{u} + \mathcal{I} p) - 2 \vg{\Omega} \times (\rho \vb{u}) - \rho \nabla \Phi,
\end{align} where $\vb{u} \otimes \vb{u}$ denotes the outer product and $\mathcal{I}$ is the identity matrix.

@@

Due to its association with hydrostatic models, it is common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.  Decomposing the momentum transport term into horizontal and vertical components gives
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w) \vb{k}.
\end{align}  The first term in this expression admits the relationships
\begin{align}
\left[ \vb{u}_h \cdot \nabla \vb{u}_h \right]_z =& \nabla_z K_2 + \zeta_h \vb{k} \times \vb{u}_h, \\
(\vb{u}_h \cdot \nabla \vb{u}_h) \cdot \vb{k} =& - \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) = - K_2 (\nabla \cdot \vb{k}) - \tfrac{1}{2} \left[ \vb{u}_h \cdot (\nabla \times \vb{u}_t)  + (\nabla \times \vb{u}_h) \cdot \vb{u}_t \right]
\end{align} where $\zeta_h = (\nabla \times \vb{u}) \cdot \vb{k} = (\nabla \times \vb{u}_h) \cdot \vb{k}$ is the relative vorticity scalar and $\vb{u}_t = \vb{k} \times \vb{u}_h$.  Note that this equation does incorporate metric terms associated with horizontal advection of $\vb{k}$ which must be accounted for.

@@

\subsection{Arbitrary vertical coordinates}

The dynamical equations are now formulated in terms of the vertical coordinate $s(t,\vb{x},z)$ with $\partial s / \partial z \neq 0$ everywhere, i.e. following \cite{Kasahara1974} (hereafter K74).  Since $\vb{x}$ and $t$ are shared between the two coordinate systems, the chain rule can be applied to obtain expressions
\begin{align} \label{eq:ChainRule}
\pdiff{}{z} =& \pdiff{s}{z} \pdiff{}{s}, & \nabla_s =& \nabla_z + (\nabla_s z) (\vb{k} \cdot \nabla), & \left( \pdiff{}{t} \right)_s =& \pdiff{}{t} + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla),
\end{align} which correspond to derivatives in the vertical, in the horizontal and in time.  This final expression is used to describe the rate of change of a quantity on $s$ surfaces.  When applied to $s$, these operators yields the useful identities
\begin{align} \label{eq:NablaZSIdentity}
\nabla_z s = - \left( \pdiff{s}{z} \right) \nabla_s z, \qquad \mbox{and} \qquad \pdiff{s}{t} = - \pdiff{s}{z} \left( \pdiff{z}{t} \right)_s.
\end{align}  From here (\ref{eq:ChainRule}) also gives rise to
\begin{align} \label{eq:HorizontalChainRule}
\nabla_z = \nabla_s - \pdiff{s}{z} (\nabla_s z) \pdiff{}{s}.
\end{align}  This equation can be used directly to rewrite (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2}) in terms of derivatives over $s$.  Note that the operators $\nabla_z$ and $\nabla_s$ are usually introduced in the context of 2D flows, however the construction described here has the advantage of working seamlessly in a 3D context, while admitting the properties $\vb{k} \cdot \nabla_z = 0$ and $\vb{k} \cdot \nabla_s = 0$.

The generalized velocity $\dot{s}$ following a fluid parcel is defined by
\begin{align} \label{eq:DefinitionDotS}
\dot{s} \equiv \diff{s}{t} = \pdiff{s}{t} + \vb{u} \cdot \nabla s = \vb{u}_h \cdot \nabla_z s + \left[ w - \left( \pdiff{z}{t} \right)_s \right] \pdiff{s}{z}.
\end{align}  Then using (\ref{eq:ChainRule}) and (\ref{eq:DefinitionDotS}) to rewrite (\ref{eq:MaterialDerivativeScalar}) gives an expression for the material derivative for scalars on $s$ surfaces,
\begin{align}
\diff{A}{t} =& \left( \pdiff{A}{t} \right)_s - \pdiff{s}{z} \left( \pdiff{z}{t} \right)_s \pdiff{A}{s} + \vb{u}_h \cdot \left[ \nabla_s A + (\nabla_z s) \pdiff{A}{s} \right] + w \pdiff{s}{z} \pdiff{A}{s} \\
=& \left( \pdiff{A}{t} \right)_s  + \vb{u}_h \cdot \nabla_s A + \dot{s} \pdiff{A}{s}  \label{eq:MaterialDerivativeS}
\end{align}

The prognostic equation for vertical velocity from (\ref{eq:VerticalVelocityOrthogonal}) is then reformulated using (\ref{eq:ChainRule}),
\begin{align} \label{eq:VerticalVelocityOrthogonalS}
\left(\pdiff{w}{t} \right)_s =& \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K + \Phi) \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k}.
\end{align}  Similarly, the prognostic equation for horizontal velocity from (\ref{eq:HorizontalVelocityOrthogonal}) is reformulated as
\begin{align} \label{eq:HorizontalVelocityOrthogonalS}
\left( \pdiff{\vb{u}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla \vb{u}_h).
\end{align}  Note that the last term of this expression can then be expanded using
\begin{align}
(\vb{k} \cdot \nabla \vb{u}_h) =& - \vb{u}_h \cdot \nabla \vb{k} - \vb{k} \times (\nabla \times \vb{u}_h),
\end{align} where under the shallow-atmosphere approximation $(\vb{u}_h \cdot \nabla \vb{k})$ is typically dropped.

The prognostic equations utilizing horizontal kinetic energy $K_2$ in place of $K$ are similarly derived from (\ref{eq:PrognosticWEquationK2}),
\begin{align} \label{eq:PrognosticWEquationK2S}
\left( \pdiff{w}{t} \right)_s =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u}_h \cdot \nabla w + \left( \pdiff{s}{z} \right) \left\{ \left[ \left( \pdiff{z}{t} \right)_s - w \right] \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K_2 + \Phi) \right\} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k},
\end{align} and from (\ref{eq:PrognosticUhEquationK2}),
\begin{align} \label{eq:PrognosticUhEquationK2S}
\left( \pdiff{\vb{u}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \vb{k} \times (\nabla \times \vb{u}_h)  - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}).
\end{align}  Observe that both of these equations are simplified if $(\partial z / \partial t)_s = w$.

{\color{red}[Is there an easy way to write down the curl (specifically relative vorticity) in arbitrary vertical coordinates?]}

@@

In conjunction with (\ref{eq:HorizontalChainRule}), the horizontal momentum equation (either  (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2})) with an arbitrary vertical coordinate gives rise to a two-term pressure gradient.  This can be avoided by prognosing the covariant components of the velocity in place of the physical velocity components.  The horizontal covariance operator is defined by
\begin{align}
[\vb{b}]_s \equiv [\vb{b}]_z + (\nabla_s z) (\vb{k} \cdot \vb{b}).
\end{align} Hence the covariant horizontal velocity is given by
\begin{align}
\vb{v}_h \equiv [\vb{u}]_s = \vb{u}_h + (\nabla_s z) w.
\end{align}  For a time-dependent $s$ coordinate,
\begin{align}
\left[ \pdiff{}{t} (\nabla_s z) \right]_s =& \nabla_s \left( \pdiff{z}{t} \right)_s,
\end{align} and so, using (\ref{eq:VerticalVelocityOrthogonalS}) and (\ref{eq:PrognosticUhEquationK2S}) the evolution equation for the covariant horizontal velocity reads
\begin{align}
\left( \pdiff{\vb{v}_h}{t} \right)_s =& \left( \pdiff{\vb{u}_h}{t} \right)_s + \left( \pdiff{w}{t} \right)_s (\nabla_s z) + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
%=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \vb{k} \times (\nabla \times \vb{u}_h)  - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}) \\
%& + (\nabla_s z) \left\{ \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K_2 + \Phi) - w \pdiff{w}{s} \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k} \right\} \\
%& + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
=& - \frac{1}{\rho} \nabla_s p - \nabla_s (K_2 + \Phi) + w \nabla_s \left( \pdiff{z}{t} \right)_s - w (\vb{u}_h \cdot \nabla \vb{k}) - [(2 \vg{\Omega} + \vg{\zeta}) \times \vb{u}]_s \nonumber \\
& \qquad - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \left[ \vb{k} \times (\nabla \times \vb{u}_h) + (\nabla_s z) \left( \pdiff{s}{z} \right) \pdiff{w}{s} \right]. \label{eq:CovariantVelocityUpdateS}
\end{align}  Again, observe that this equation simplifies significantly if $(\partial z / \partial t)_s = w$.

%For a time-dependent $s$ coordinate,
%\begin{align}
%\pdiff{}{t} (\nabla_s z) =& \pdiff{}{t} \left[ - (\nabla_z s) \left( \pdiff{s}{z} \right)^{-1} \right] \\
%=& - \left( \nabla_z \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} + (\nabla_z s) \left( \pdiff{s}{z} \right)^{-2} \pdiff{}{z} \pdiff{s}{t} \\
%=& - \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right)
%\end{align}
%For a time-dependent $s$ coordinate,
%\begin{align}
%\left[ \pdiff{}{t} (\nabla_s z) \right]_s =& - \left( \nabla_z \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} + (\nabla_z s) \left( \pdiff{s}{z} \right)^{-2} \pdiff{}{z} \pdiff{s}{t} + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla) (\nabla_s z) \\
%=& - \left( \nabla_z \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} + (\nabla_z s) \left( \pdiff{s}{z} \right)^{-2} \pdiff{}{z} \pdiff{s}{t} + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla) \left[ - (\nabla_z s) \left( \pdiff{s}{z} \right)^{-1} \right] \\
%=& - \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right) - \left( \pdiff{z}{t} \right)_s \left( \pdiff{s}{z} \right)^{-1} \nabla_z \left( \pdiff{s}{z} \right) + \left( \pdiff{z}{t} \right)_s \left( \pdiff{s}{z} \right)^{-2} (\nabla_z s) \pdiff{}{z} \pdiff{s}{z} \\
%=& \left( \pdiff{s}{z} \right)^{-1} \left[ - \nabla_s \left( \pdiff{s}{t} \right) - \left( \pdiff{z}{t} \right)_s \nabla_s \left( \pdiff{s}{z} \right)  \right] \\
%=& \nabla_s \left( \pdiff{z}{t} \right)_s
%\end{align}

%Then using (\ref{eq:MaterialDerivativeS}) gives
%\begin{align}
%\left( \pdiff{\vb{v}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_s p - \nabla_s (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_s - w \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right) - \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla \vb{v}_h) 
%\end{align}

%Although a time-dependent $s$ is admitted by this equation, its fully Eulerian nature makes it difficult to use in this case since degrees of freedom are not guaranteed to align with constant $s$ surfaces.

%\subsection{Floating Lagrangian Formulation}
